Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-23 paper.tex
Fix line ends

\section{Introduction}
% Recombination is an important force in SARS-CoV-2, recombinants
% have arisen and they have spread to high frequencies
Recombination via template switching is a common feature
in the evolution of coronaviruses~\citep{Graham2010-xe,De_Klerk2022-tt},
including SARS-CoV-2
\citep{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing
together mutations carried by different lineages, recombination plays an
important role in generating genomic diversity, with recombinant lineages
associated with adaptation to new host species and with the production of more
immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.
Early in the SARS-CoV-2 pandemic there was insufficient 
genetic variation for meaningful recombination.
In late 2020, the appearance and
spread of variants of concern (VoC), designated into classes such as Alpha and
Delta which harboured multiple characteristic mutations,
created the conditions required for recombination to occur
and recombinant strains with onward transmission were detected at
low frequency~\citep{Jackson2021-ik}.
More recently, the high prevalence of Omicron,
with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has
accelerated the rate of coinfection and the potential for
recombination~\citep{Bal2022-hq}.
In early 2023, multiple recombinant lineages have successfully
established and spread to high frequency, and accounting for recombinant
ancestry is now essential in understanding the evolution of the SARS-CoV-2
virus.

% Recombination is hard to detect
Detecting recombination in SARS-CoV-2 is difficult, and identifying new
recombinant strains is a time-consuming, manual process, subject 
to biases [FIXME: have we a citation for how tedious this is? Also what 
are the biases?]
Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
combine specific subsets of mutations from different lineages
\citep[e.g.][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz},
which can only identify inter-lineage recombination events specific to those motifs.
% % JK: I've left this out because I felt it broke the narrative up a bit, but
% % happy to put back in. (We're still citing the paper in a few places).
% Others have inferred recombinant
% genealogies to disentangle recombination from recurrent mutation, without
% having to pre-specify characteristic mutations for each lineage
% \citep{Ignatieva2021-rg}.
\cite{Turakhia2022-it} presented a
phylogeny-based approach (``RIPPLES'') to identify putative recombinants among
over ten million SARS-CoV-2 genomes, without pre-specifying motifs. 
RIPPLES finds candidate recombinants by using an existing phylogeny
(built assuming no recombination), and finding potential recombinants
by  scanning for branches containing many mutations.
It then determines if these candidates would be better explained by
recombination by exhaustively breaking each 
sequence into segments and attempting to find more parsimonious placements for
each segment on the phylogeny. If such placements are found, the sequence is
identified as a putative recombinant. Although it enables rapid searches for
genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2 phylogeny that
accounts for only mutations, treats recombinants \textit{post hoc}, and is an
incomplete representation of the reticulate evolutionary history of SARS-CoV-2.
As noted by the authors, a \textit{post hoc} treatment of recombination is
possible when recombinant lineages are rare and leave few descendants. However,
the proliferation of recombinant lineages is making this increasingly
untenable; for example, more than half of the sequences sampled in February 2023
are from XBB and its descendants~\citep{Chen2022-pz}. This also means that
future evolution of SARS-CoV-2 is likely to involve multiple sequential
recombination events on top of existing recombinant lineages, creating a highly
reticulated ancestry.

@@

It is well known that recombination distorts phylogenies \citep{Schierup2000-fg}
and affects the results of downstream analysis, e.g., inference of
selection~\citep{Anisimova2003-vr}. Standard phylogenetic methods do
not account for recombination
\citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}, and there is
no standard method for incorporating the effects of recombination into
phylogenetic analysis.
Ancestral Recombination Graphs (ARGs) are a means of describing such
network-like ancestry~\citep{Griffiths1981-lw,Gusfield2014-qw}, but
until recently lacked software support and sufficiently scalable
inference methods to be of practical use.
However, approaches to infer ARGs now exist that can scale to tens of
thousands of human genomes and beyond
\citep{Speidel2019-yh,Kelleher2019-ba,Schaefer2021-yg,Zhang2023-lf}, and which
deal with levels of repeated recombination far in excess of those seen in viral
phylogenies. The ``succinct tree sequence'' is an ARG data structure
which has led to significant computational advances across a range
of
applications~\citep{Kelleher2016-wk,Kelleher2018-xc,Kelleher2019-ba,Ralph2020-efficiently,
Wohns2022-th}, and the supporting \texttt{tskit} software library 
is now widely used in population genetic applications.
The methods in \texttt{tskit} have been developed to support millions of
whole human genomes~\citep{Kelleher2019-ba}, and so it is particularly well suited
to representing large SARS-CoV-2 genealogies,
which currently encompasses over 15 million sequences in the GISAID database
\citep{Shu2017-hp}.
See Section~\ref{sec:args} for more details on ARGs and the succinct
tree sequence data structure.

% [PARA 4: we present sc2ts based on these recent advances. Summarise
% results, with forward refs to Results sections.]
Here we present a preliminary version of \texttt{sc2ts}, a novel method
for inferring ARGs for SARS-CoV-2 at pandemic scale, in real-time.
Building on the open-source \texttt{tskit} library, the method infers
a joint model of mutation, genetic inheritance and recombination, which
can be conveniently and efficiently analysed using standard data science
tools. Inference is based on incrementally adding batches of sequences
based on their collection dates, and proceeds in three phases. 
First, likely attachment nodes in the current ARG
(or paths, if inferred to be a recombinant) are
chosen for each of the samples in the daily batch using the
Li and Stephens model (Section~\ref{sec:ls}). Then, since many samples
in a daily batch can share an attachment path,
we infer phylogenetic trees for each of the clusters separately using standard
methods (Section~\ref{sec:sample-cluster-tree-inference}).
Finally, we attach the trees for these sample-clusters to the current
ARG, and apply some parsimony heuristics to address issues
introduced by the inherent greediness of the strategy
(Section~\ref{sec:parsimony-heuristics}).
Using the current preliminary version of \texttt{sc2ts} we infer two
large ARGs (1,265,685 and 657,239 samples, respectively), and study
the properties of these ARGs to illustrate the power of the method
and to inform subsequent development. We find that these
ARGs accurately reflect known patterns of phylogenetic relationships
between samples (Section~\ref{sec:backbone_phylogeny}),
mutational spectra (Section~\ref{sec:mutation_spectrum}),
and automatically identify the majority of known recombinant
lineages (Sections~\ref{sec:jackson_recombs} and \ref{sec:pango_x_lineages}),
with an unprecedented degree of precision in the
genomic location of recombination breakpoints
(Section~\ref{sec:breakpoint_intervals})
and the relationship between parental sequences
(Section~\ref{sec:parent_divergence}).
We hope that these benefits of an accurate joint model of
mutation, genetic inheritance and recombination, will generate community
interest and development of the \texttt{sc2ts} method, and
more generally in applying the efficient and mature software
of the \texttt{tskit} ecosystem to pandemic-scale SARS-CoV-2 data.

@@

% Paragraph 2, what do we plot, specifically?
In this section we investigate the patterns of divergence between
the parents of recombination nodes.
We focus on the Long ARG because it covers time
periods where recombination is known to have occurred
a large number of times, for example it contains samples from over
30 Pango X lineages (i.e., those inferred
to have recombinant ancestry; see Section~\ref{sec:pango_x_lineages} for
further analysis of Pango recombinant lineages in the Long and Wide ARGs).
Figure~\ref{fig:recomb_mrcas} shows the estimated date of the most recent
common ancestor (MRCA) of the parent nodes for each recombination breakpoint,
plotted against the divergence between these parents (i.e. the
total branch length from the parents to their MRCA in the trees to
the immediate left and right of the breakpoint).
As in Section~\ref{sec:breakpoint_intervals}, in these plots we
exclude ``singleton'' recombination nodes (those ancestral to only one sample).
Figure~\ref{fig:recomb_mrcas}A plots breakpoints for all nonsingleton recombinants,
and Figure~\ref{fig:recomb_mrcas}B plots a subset of potentially more
reliable breakpoints that occur in nodes
ancestral to 10 or more samples (note that this is one of the official criteria
used when proposing Pango recombinant lineages [CITE]).
% that we deem to be more reliable by the measure of ``HMM consistency''. 
% We define an  in that when we compare the LS HMM results when run from
% 5' to 3' versus 3' to 5', it gives the
% same mutations patterns and sequence-identical 5' parents as when run 3' to 5'
This reduces the number of recombination nodes
to XXX and breakpoints to YYY.
Note that each plotted point
represents a breakpoint, so that a recombination node with more than one
breakpoint (e.g. with 3 or more parents, comprising ${\sim}10 \%$ of the
recombination nodes in Figure ~\ref{fig:recomb_mrcas}A and ${\sim}3 \%$ in
Figure ~\ref{fig:recomb_mrcas}B) will be represented by several points.
It is important to note here that some of the time estimates in these figures
come with a substantial degree of uncertainly,
since nodes other than samples are dated somewhat crudely in the current version
of \texttt{sc2ts} (Section~\ref{sec:node_dating}). Nevertheless, the histograms
clearly show that \texttt{sc2ts} can identify recombination between lineages
that are only a matter of weeks apart.

>>>>

\section{Introduction}
% Recombination is an important force in SARS-CoV-2, recombinants
% have arisen and they have spread to high frequencies
Recombination via template switching is a common feature
in the evolution of coronaviruses~\citep{Graham2010-xe,De_Klerk2022-tt},
including SARS-CoV-2
\citep{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing
together mutations carried by different lineages, recombination plays an
important role in generating genomic diversity, with recombinant lineages
associated with adaptation to new host species and with the production of more
immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.
% Early in the SARS-CoV-2 pandemic there was insufficient 
% genetic variation for meaningful recombination.
Early on in the COVID-19 pandemic, the levels of genetic diversity
were too low for the emergence of distinctive recombinant strains.
By late 2020, however, the appearance and
spread of variants of concern (VoC), designated into classes such as Alpha and
Delta which harboured multiple characteristic mutations,
created the conditions required for recombination to occur
and recombinant strains with onward transmission were detected at
low frequency~\citep{Jackson2021-ik}.
More recently, the high prevalence of Omicron,
with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has
accelerated the rate of coinfection and the potential for
recombination~\citep{Bal2022-hq}.
In early 2023, multiple recombinant lineages have successfully
established and spread to high frequency, and accounting for recombinant
ancestry is now essential in understanding the evolution of the SARS-CoV-2
virus.

% Recombination is hard to detect
Detecting recombination in SARS-CoV-2 is difficult, and identifying new
recombinant strains is a time-consuming, manual process, subject 
to biases [FIXME: have we a citation for how tedious this is? Also what 
are the biases?]
Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
combine specific subsets of characteristic mutations from different lineages
\citep[e.g.][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz},
and as a result can only identify inter-lineage recombination events.
% specific to those motifs.
% % JK: I've left this out because I felt it broke the narrative up a bit, but
% % happy to put back in. (We're still citing the paper in a few places).
% Others have inferred recombinant
% genealogies to disentangle recombination from recurrent mutation, without
% having to pre-specify characteristic mutations for each lineage
% \citep{Ignatieva2021-rg}.
\cite{Turakhia2022-it} presented a
phylogeny-based approach (``RIPPLES'') to identify putative recombinants among
over ten million SARS-CoV-2 genomes, without pre-specifying sets of characteristic
mutations.
RIPPLES finds candidate recombinants by using an existing phylogeny
(built assuming no recombination), and finding potential recombinants
by  scanning for branches containing many mutations.
It then determines if these candidates would be better explained by
recombination by exhaustively breaking each 
sequence into segments and attempting to find more parsimonious placements for
each segment on the phylogeny. If such placements are found, the sequence is
identified as a putative recombinant. Although it enables rapid searches for
genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2 phylogeny that
accounts for only mutations, treats recombinants \textit{post hoc}, and is an
incomplete representation of the reticulate evolutionary history of SARS-CoV-2.
As noted by the authors, a \textit{post hoc} treatment of recombination is
possible when recombinant lineages are rare and leave few descendants. However,
the proliferation of recombinant lineages is making this increasingly
untenable; for example, more than half of the sequences sampled in February 2023
are from XBB and its descendants~\citep{Chen2022-pz}. This also means that
future evolution of SARS-CoV-2 is likely to involve multiple sequential
recombination events on top of existing recombinant lineages, creating a highly
reticulated ancestry.

@@

It is well known that recombination distorts phylogenies \citep{Schierup2000-fg}
and affects the results of downstream analyses, e.g., inference of
selection~\citep{Anisimova2003-vr}. Standard phylogenetic methods do
not account for recombination
\citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}, and there is
no standard method for incorporating the effects of recombination into
phylogenetic analysis.
Ancestral Recombination Graphs (ARGs) are a means of describing such
network-like ancestry~\citep{Griffiths1981-lw,Gusfield2014-qw}, but
until recently lacked software support and sufficiently scalable
inference methods to be of practical use.
However, approaches to infer ARGs now exist that can scale to tens of
thousands of human genomes and beyond
\citep{Speidel2019-yh,Kelleher2019-ba,Schaefer2021-yg,Zhang2023-lf}, and which
deal with levels of repeated recombination far in excess of those seen in viral
phylogenies. The ``succinct tree sequence'' is an ARG data structure
which has led to significant computational advances across a range
of
applications~\citep{Kelleher2016-wk,Kelleher2018-xc,Kelleher2019-ba,Ralph2020-efficiently,
Wohns2022-th}, and the supporting \texttt{tskit} software library 
is now widely used in population genetic applications.
The methods in \texttt{tskit} have been developed to support millions of
whole human genomes~\citep{Kelleher2019-ba}, and so it is particularly well suited
to representing large SARS-CoV-2 genealogies,
which currently encompasses over 15 million sequences in the GISAID database
\citep{Shu2017-hp}.
See Section~\ref{sec:args} for more details on ARGs and the succinct
tree sequence data structure.

% [PARA 4: we present sc2ts based on these recent advances. Summarise
% results, with forward refs to Results sections.]
Here we present a preliminary version of \texttt{sc2ts}, a novel method
for inferring ARGs for SARS-CoV-2 at pandemic scale, in real-time.
Building on the open-source \texttt{tskit} library, the method infers
a joint model of mutation, genetic inheritance and recombination, which
can be conveniently and efficiently analysed using standard data science
tools. Inference is based on incrementally adding batches of sequences
based on their collection dates, and proceeds in three phases. 
First, likely attachment nodes in the current ARG
(or paths, if inferred to be a recombinant) are
chosen for each of the samples in the daily batch using the
Li and Stephens model (Section~\ref{sec:ls}). Then, since many samples
in a daily batch can share an attachment path,
we infer phylogenetic trees for each of the clusters separately using standard
methods (Section~\ref{sec:sample-cluster-tree-inference}).
Finally, we attach the trees for these sample-clusters to the current
ARG, and apply some parsimony heuristics to address issues
introduced by the inherent greediness of the strategy
(Section~\ref{sec:parsimony-heuristics}).
Using the current preliminary version of \texttt{sc2ts}, we infer two
large ARGs (1,265,685 and 657,239 samples, respectively), and study
the properties of these ARGs to illustrate the power of the method
and to inform subsequent development. We find that these
ARGs accurately reflect known patterns of phylogenetic relationships
between samples (Section~\ref{sec:backbone_phylogeny}),
mutational spectra (Section~\ref{sec:mutation_spectrum}),
and automatically identify the majority of known recombinant
lineages (Sections~\ref{sec:jackson_recombs} and \ref{sec:pango_x_lineages}),
with an unprecedented degree of precision in the
genomic location of recombination breakpoints
(Section~\ref{sec:breakpoint_intervals})
and the relationship between parental sequences
(Section~\ref{sec:parent_divergence}).
We hope that these benefits of an accurate joint model of
mutation, genetic inheritance and recombination, will generate community
interest and development of the \texttt{sc2ts} method, and
more generally in applying the efficient and mature software
of the \texttt{tskit} ecosystem to pandemic-scale SARS-CoV-2 data.

@@

% Paragraph 2, what do we plot, specifically?
In this section we investigate the patterns of divergence between
the parents of recombination nodes.
We focus on the Long ARG because it covers time
periods where recombination is known to have occurred
a large number of times, for example it contains samples from over
30 Pango X lineages (i.e., those inferred
to have recombinant ancestry; see Section~\ref{sec:pango_x_lineages} for
further analysis of Pango recombinant lineages in the Long and Wide ARGs).
Figure~\ref{fig:recomb_mrcas} shows the estimated date of the most recent
common ancestor (MRCA) of the parent nodes for each recombination breakpoint,
plotted against the divergence between these parents (i.e. the
total branch length from the parents to their MRCA in the trees to
the immediate left and right of the breakpoint).
As in Section~\ref{sec:breakpoint_intervals}, in these plots we
exclude ``singleton'' recombination nodes (those ancestral to only one sample).
Figure~\ref{fig:recomb_mrcas}A plots breakpoints for all nonsingleton recombinants,
and Figure~\ref{fig:recomb_mrcas}B plots a subset of potentially more
reliable breakpoints that occur in nodes
ancestral to 10 or more samples (note that this is one of the official criteria
used when proposing Pango recombinant lineages [CITE]).
% that we deem to be more reliable by the measure of ``HMM consistency''. 
% We define an  in that when we compare the LS HMM results when run from
% 5' to 3' versus 3' to 5', it gives the
% same mutations patterns and sequence-identical 5' parents as when run 3' to 5'
This reduces the number of recombination nodes
to 161 and breakpoints to 174.
Note that each plotted point
represents a breakpoint, so that a recombination node with more than one
breakpoint (e.g. with 3 or more parents, comprising ${\sim}10 \%$ of the
recombination nodes in Figure ~\ref{fig:recomb_mrcas}A and ${\sim}3 \%$ in
Figure ~\ref{fig:recomb_mrcas}B) will be represented by several points.
We caution that some of the time estimates in these figures
come with a substantial degree of uncertainly,
since non-sample nodes are dated somewhat crudely in the current version
of \texttt{sc2ts} (Section~\ref{sec:node_dating}). Nevertheless, the histograms
clearly show that \texttt{sc2ts} can identify recombination between lineages
that are only a matter of weeks apart.
